---
title: "IPRT Survey - Educational Attainment"
author: "Cormac Nolan"
email: "cormacnolan85@gmail.com"
output:
  html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
  html_notebook: default
  pdf_document: default

  
---


```{r setup, echo = F, include = F}
# If you're new to R, this RMarkdown document is a combination between chunks of R code and standard markdown, much like a Jupyter notebook. The header above simply defines what kind of output to make, whether to make a TOC etc.

# This setup code chunk simply sets some 

library(tidyverse)
# Tidyverse is a wrapper package for a suite of *essential* R packages that
# basically acts like a completely different API inside R, who's base API
# is famously/notoriously concise/impenetrable.
#
# The constituent packages were primarily conceived and built over years 
# by a complete legend of a New Zealander called Hadly Wickham, who is 
# as close to a celbrity as you can get among stats programmers.
# I highly (HIGHLY!) recommend using this API as your starting point in
# R if you're new to it.
#
# The tidyverse includes:
#   - dplyr for data pre-processing & cleaning
#   - ggplot2 for general plotting, far superior to the base R graphics.
#   - readr for importing data conveniently
#   - tidyr for putting data into "tidy" form. Often used in combo with dplyr.
#   - etc
#   - get the full list with: tidyverse::tidyverse_packages()

# Knitr is used to create documents from this RMarkdown.
# I'm using it here to set default options for how this document will "knit"
library(knitr) 


opts_chunk$set(fig.width=8, fig.height=8)
opts_chunk$set(echo = F, warning = F, error = F, message = F)

# This just runs another file which puts a consistent palette and
# ggplot2 theme into memory for later plotting.
source('../Utilities/create_theme.R')
```
#Introduction
This [RMarkdown](http://rmarkdown.rstudio.com/)-generated document is intended to describe and quantify the difference in attainment in education between the general population, as quantified in the Irish 2011 census, and the prison population that is quantified in the IPRT survey. 

The document is a "living document" of my attempts to understand and analyse the data-set and as you progress through the document you'll see the refinement of the analysis from initial exploration plots, to further data cleaning, statistical analysis and finally the last plots and summaries that quantify the differences between the populations in a usable and understandable way.

# Executive Summary
* There is a large and statistically significant difference in educational attainment between the prison population and the general population, as described in the Irish 2011 census. 
* There is ahuge  difference in the proportion of inmates that have no education or only primary education, when compared to the general population. That's ~50% in the former and ~16 % in the latter. The next largest category among inmates is lower secondary completed only. 
* Those who have only lower secondary or less, make up ~74% of the prison population, compared to ~34% in the general population.
* When looking at the educational attainment within each age group, the dearth of data points at ages > 50 years meant conclusions could not be drawn for the older inmate population.
* The proportion who have completed some form of 3rd level education goes up as the age increases above 36 among inmates. What's going on here is impossible to tell from the current data, but it is interesting that this pattern doesn't seem to appear in the general population.
* The IPRT survey used education categories that were different from the census' education categories, which were different again to international standards like ISCED. This meant that certain data points had to be merged into broader categories and lead to a subsequent loss of resolution in the analysis. Similarly for the age categories used in the survey.
* Please see section 8.3 for the most "useful" plot in visualising the difference betwen the general and prison population by age category.
* For a view of the difference between the two populations overall, see section 6.2

#Import & Clearning
After setup, first thing is to get the survey results loaded and do a bit of filtering & cleaning in order to get the data we're concerned with, i.e., educational data plus some other demographic data for later.

Here I'm using the Tidyverse set of apps, including readr and dplyr.
```{r read & clean}

# First i'm using readr to import the survey data
survey <- 
  readr::read_csv("../../../Data/survey_results_clean.csv")

# Now I'm using dplyr to clean the data up a bit.
# %>% is a pipe, that puts the output of one function as the first 
# argument of the next function, much like a Unix pipe. It is VERY handy.
survey_education <-
  survey %>% 
  select(X1, gender, age, school_experience, # select the relevant columns 
         prison_wing, prison, school_experience_level) %>% 
  mutate(gender = as.factor(gender), # mutate changes columns or creates new ones
         age = as.factor(age),
         school_experience = 
           ordered(school_experience, 
                   levels = c("I didnt go to secondary school.",
                              "I left school before the Junior/Inter Cert.",
                              "I left school after the Junior/ Inter Cert.",
                              "I left school after the Leaving Cert.",
                              'I have gone to college/ university.',
                              "Unclear")), # google "R factors" to understand 
         prison_wing = as.factor(prison_wing),
         prison = as.factor(prison),
         school_experience_level = as.factor(school_experience_level))



```

# Exploration Plots
First lets do some basic plots to explore our data:

```{r explore survey}

# Our first plot using ggplot2. The syntax is a bit weird before you get 
# used to it, but basically we put our data frame as the first argument
# and the second argument is an aesthetic (aes) that tells ggplot which
# columns to use in plotting data and also how to colour encode.
# In this case we want to plot the school_experience_level and we use
# fill signify that we want the bars to be filled with 
# colours based on on the school_experience variable.
# We then build our plot, layer by layer with the + operator used to 
# add the layers or modify previous layers.
# Geoms are the main "controllers" of the type of plot, in this case a bar.
# The iprt_theme is a pre-baked theme I imported earlier and
# the iprt_palette is used to dedice the colour of the fill in the bar plot.

ggplot(survey_education, aes(school_experience_level, 
                             fill = school_experience)) +
  geom_bar() + # automatically sums up over the bins to create the bars
  iprt_theme +
  scale_fill_manual(values = iprt_palette[-1])



```

It seems the population is (as expected) skewed towards the lower end of educational attainment. This will become very obvious in the next plot, showing similar data for the general population.

```{r explore general population}

# Data retrieved from: http://www.cso.ie/en/statistics/education/principalstatistics/
# Date: 2011
# Retrieved: 19/02/2017

# Now i'm importing the basic census data on educational attainment.
# Notice how I use the pipes to add a new variable in the same statement
# as the import. This can make the code far more concise & natural
general_pop_education <-
  read_csv("../../../Data/External Data/educational_attainment_2011.csv") %>% 
  mutate(school_experience_level = c(NA, NA, 1, 2, 3, 4, 5, 6, NA, NA))
  
# A similar plot to above.
ggplot(general_pop_education %>% filter(category_type == "ceased"),
       aes(x = reorder(category, school_experience_level), y = male,
           fill = category)) +
  geom_bar(stat="identity") + #stat="identity" is used since we have the totals
  iprt_theme +
  scale_fill_manual(values = iprt_palette[-1])


```

#CSO vs. Survey
Unfortunately the CSO uses a different metric for educational attainment as the IPRT survey, so I used an international standard (ISCED 2011) and mapped our other data-sets to that and then inserted this mapping variable into the two data-sets.

```{r map between general and survey}

# Using mutate to create a new variable to map the census data to ISCED
general_pop_education <-
  general_pop_education %>% 
  mutate(ISCED_2011 = c(NA, NA, "0-1", "2", "3", 
                        "4-8", "4-8", "unknown",
                        NA, NA))

# Similarly for the survey data but using case_when() to define the map
survey_education <- 
  survey_education %>% 
  mutate(ISCED_2011 = 
           case_when(.$school_experience_level %in% c("A", "B") ~ "0-1",
                     .$school_experience_level == "C" ~ "2",
                     .$school_experience_level == "D" ~ "3",
                     .$school_experience_level == "E" ~ "4-8",
                     .$school_experience_level == "Unclear" ~ "unknown"))
# Btw the $ operator accesses the columns of a data frame, and the . (dot)
# is used to reference the data frame that has been passed in by the pipe %>%
```

There are two downsides that I can see here, one is that we lose granularity in our categories and the second is that there is a certain amount of ambiguity around the mapping (for example does ISCED 2011 category 4 really map correctly to "I have gone to college/ university" in the survey).

The upside is that we now have a common metric with which we can compare the two data-sets.

#Inmates vs General Population: ISCED 2011
Using the [ISCED 2011](https://en.wikipedia.org/wiki/International_Standard_Classification_of_Education#ISCED_2011_levels_of_education_and_comparison_with_ISCED_1997) we can redo the previous plots with a common metric and the difference between the populations is immediately obvious. Far fewer inmates are completing secondary school or going on to 3rd level than in the general population, based on the two samples we have. The problem here is that the inmate population might be significantly skewed compared to the general population so we may have to adjust for that to arrive at a proper conclusion.

## ISCED Quick Guide
0 = No primary education
1 = Completed primary education
2 = Completed lower secondary education
3 = Completed upper secondary education
4-8 = Some form of college or post-secondary education, all the way up to PhD level.

## Plots Comparing Attainment
```{r ISCED 2011 Inmates}
 
ggplot(survey_education, aes(ISCED_2011, 
                             fill = ISCED_2011)) +
  geom_bar() +
  iprt_theme +
  scale_fill_manual(values = iprt_palette[-1]) +
  labs(title = "Educational Attainment of Inmates Using\nISCED 2011 Standard",
       x = "ISCED 2011 Level of Attainment", y = "Inmates") 
```

```{r ISCED 2011 General}

# here I'm using dplyr to filter (like the "WHERE" in SQL) the data as I plot it.
# This means I don't need to keep multiple similar copies of a dataset.
ggplot(general_pop_education %>% filter(category_type == "ceased"),
       aes(x = ISCED_2011, y = male,
           fill = ISCED_2011)) +
  geom_bar(stat="identity") +
  iprt_theme +
  scale_fill_manual(values = iprt_palette[-1]) +
  labs(title = 
         "Educational Attainment of Male General Population\nUsing ISCED 2011 Standard",
       x = "ISCED 2011 Level of Attainment", y = "Male General Population")
```

#Normalisation and Statistical Comparison

```{r survey education summary}

# More dplyr, this time to create summaries based on groups.
# group_by is similar to the SQL statement but only defines the group,
# what you do with it comes in the subsequent statements which are usually
# a tally() or summarise().

survey_education_summary <-
  survey_education %>% 
  group_by(ISCED_2011) %>% 
  tally() %>% # does a sum across each category
  mutate(proportion = n/sum(n)) # use mutate to create a new proportion variable

survey_education_summary

survey_education_summary$n %>% sum()
  
```

```{r general education summary}
general_education_summary <-
  general_pop_education  %>% 
  filter(category_type == "ceased") %>% 
  group_by(ISCED_2011) %>% 
  # this time we use summarise as we already have totals
  summarise(n = sum(male)) %>%  
  mutate(proportion = n / sum(n))


general_education_summary

general_education_summary$n %>%  sum()

```


##Parametric Z-Test
The following shows the p-value for the test of significant difference between the two populations' for each ISCED category. Each of them is well below the standard cutoff of 0.05 leading us to reject the null hypothesis, that there is no difference between the categories.
```{r Parametric Z-Test}

# First create an empty list to which we'll insert an arbitrary results object.
# This pattern is quite commonly used in R, with a large set of  objects 
# it is often best to pre-define the length of this list because R is bad at 
# memory, but in this case we can get away with it.
ztest_results = list() 

# Here I'm running a proportional z-test on each category between the 
# survey and census data and storing the results in ztest_results.
# I'm also printing a human-readable output.
# This should probably be captured in a function or two...
for (i in 1:nrow(general_education_summary)){
  # [] in R means you're using indexed accessing of an object.
  # [[]] means you don't want the list element itself but that you want
  # to access the inner object. Otherwise you'll often get the 
  # "toString()" equivalent of the object rather than the object itself.
  ztest_results[[i]] <-
    (prop.test(x = c(general_education_summary$n[i],
                     survey_education_summary$n[i]),
               n = c(sum(general_education_summary$n),
                     sum(survey_education_summary$n)),
               correct = FALSE))
  
  paste("P Value for ISCED categoy:", 
        general_education_summary$ISCED_2011[i],
        "    ",
        ztest_results[[i]]$p.value) %>% # notce the [[]] notation again
    print()
  
}

# The strength of this approach is we can then access the entire results object
# at a later point with ease.


```


# Controlling for Age
Looking at the other work on age distributions it does look like there is a skew towards the younger ages in the prison population. In order to account for this we'll need to  import census educational attainment & age data and compare to the prison population. Statbank CD901 - for 2011 Census, 2016 not available as far as I could see.

## Import Census Age & Education
As we can see below, the categories don't match up, much like the education categories. The next step is to create a mapping between them.
```{r}

# Next I found the educational attainment census data broken down by age.
# It wasn't quite in "tidy" form, the age variable was split out among multiple
# columns,  so i used gather to create a single variable of age and a key 
# variable created from the names of the columns that I'm "gathering" into
# a key-value pairing of columns.

general_age_education_raw <- 
  read_csv("../../../Data/External Data/general_population_age_educational_attainment.csv") %>% 
  rename(educational_attainment = X1) %>% 
  tidyr::gather(age, general_pop_count, c(2:16)) %>%
  select(-2)

# Just taking a look at the distinct values of the age categories in the
# two datasets.
survey_education %>% 
  distinct(age) %>% # returns only distinct values
  arrange(age) # arrange sorts them alphanumerically.

general_age_education_raw %>% 
  distinct(age) %>% 
  arrange(age)
  

```


## Age and Education Mapping
Mapping to common age and education categories, we find that there aren't really enough samples to say much among over 50's in the prison population and the value in lumping them into a younger age category seems dubious to me. Otherwise the results bare out as before as there is a large skew towards a lower attainment in education. 
```{r}

# Create new variables to map the educational attainment and age for 
# the census data. The mappings don't match up exactly unfortunately.
#
# Note that the 15-19 age category has been filtered from the census 
# data as these can't really be compared with the prison population which
# are all 18+.

# This next couple of dplyr chunks brings what I did before all into 
# one place, so take some time to understand what's going on here.
general_age_education <-
  general_age_education_raw %>% 
  mutate(ISCED_2011 = 
           case_when(.$educational_attainment ==
                       "Primary (incl. no formal education)" ~ "0-1",
                     .$educational_attainment == "Lower secondary" ~ "2",
                     .$educational_attainment == "Upper secondary" ~ "3",
                     .$educational_attainment %in% 
                       c("Third level non-degree",
                         "Third level degree or higher") ~ "4-8",
                     .$educational_attainment == "Not stated" ~ "unknown"))%>% 
  filter(age != '15 - 19 years') %>% 
  mutate(common_age_category = 
           case_when(.$age == "20 - 24 years" ~ "< 26",
                     .$age %in% c("25 - 29 years",
                                  "30 - 34 years") ~ "26-35",
                     .$age %in%  c("35 - 39 years",
                                   "40 - 44 years",
                                   "45 - 49 years") ~ "36-49",
                     .$age %in%  c("50 - 54 years",
                                   "55 - 59 years",
                                   "60 - 64 years") ~ "50-64",
                     .$age %in%  c("65 - 69 years",
                                   "70 - 74 years",
                                   "75 - 79 years",
                                   "80 - 84 years",
                                   "85 years and over") ~ "65+")) %>% 
  group_by(common_age_category, ISCED_2011) %>% 
  summarise(general_pop_count = sum(general_pop_count)) 

survey_age_education <-
  survey_education %>% 
  mutate(common_age_category = 
           case_when(.$age %in%
                       c("18-21", "22-25") ~ "< 26",
                     .$age == "26-35" ~ "26-35",
                     .$age == "36-49" ~ "36-49",
                     .$age == "50-64" ~ "50-64",
                     .$age == "65+" ~ "65+")) %>%
  group_by(common_age_category, ISCED_2011) %>% 
  tally() %>% 
  rename(prison_pop_count = n)

combined_age_education <-
  survey_age_education %>% 
  # left join does exactly what you think it does. Really useful statement.
  left_join(general_age_education,
            by = c("common_age_category", "ISCED_2011")) %>% 
  tidyr::gather(pop_type, pop_count, c(3:4))
# We used gather() as again we had a variable (count) spread out in 2 
# columns (survey count and census count), so we combined them
# into a key column (census or survey) and value column (count).

print("Combined Prison & General Population Dataset:")
combined_age_education


```


```{r}

ggplot(general_age_education %>% 
         filter(!(common_age_category %in% c("50-64","65+"))),
       aes(x = ISCED_2011, y = general_pop_count,
           fill = ISCED_2011)) +
  geom_bar(stat="identity") +
  # facets split the plots into a multiplot based on the variable you pass in
  facet_wrap(~common_age_category) + 
  iprt_theme +
  scale_fill_manual(values = iprt_palette[-1]) +
  labs(title = 
         "Educational Attainment of Male General\nPopulation by Age (2011)",
       x = "ISCED 2011 Level of Attainment", y = "Male General Population")

ggplot(survey_age_education %>% 
         filter(!(common_age_category %in% c("50-64","65+"))),
       aes(x = ISCED_2011, y = prison_pop_count,
           fill = ISCED_2011)) +
  geom_bar(stat="identity") +
  facet_wrap(~common_age_category) +
  iprt_theme +
  scale_fill_manual(values = iprt_palette[-1]) +
  labs(title = 
         "Educational Attainment of Prison Population by Age",
       x = "ISCED 2011 Level of Attainment", y = "Prison Population")



```

## Normalised Comparison
If we normalise each category to facilitate comparison with the general population, and focus on the ages then things become more clear. We can see below just how much of a difference there is in each age category. This means we don't have to control for age, as we are comparing like with like and have reasonable sample sizes. Again, the 50+ age categories have been left out due to a small sample.

It should be noted that due to the sample size on the census, the variance in these categories is tiny compared to what is likely to be the variance in the prison population results.

```{r}

ggplot(combined_age_education %>% 
         filter(!(common_age_category %in% c("50-64","65+"))),
       aes(x = common_age_category, y = pop_count,
           fill = ISCED_2011)) +
  geom_bar(stat="identity", position = "fill") +
  iprt_theme +
  scale_fill_manual(values = iprt_palette[-1]) +
  facet_wrap(~pop_type) +
  labs(title = 
         "Educational Attainment of Prison &\nGeneral Population by Age",
       x = "ISCED 2011 Level of Attainment", y = "Proportion")

```

What's most striking to me is the order of magnitude difference in the proportion of inmates that have no education or only primary education, when compared to the general population. That's ~50% in the former and ~5% in the latter. The next largest category among inmates is lower secondary completed only, together these make up ~75-80% of prison population.

What is also interesting is that the proportion who have completed some form of 3rd level education goes up as the age increases above 36 among inmates. What's going on here is impossible to tell from the current data, but it is interesting that this pattern doesn't seem to appear in the general population. 

```{r}

# Here I'm replicating the calculations done by ggplot to make the above plots
# in dplyr. It's quite straight-forward to do when you get the hang of it 
# and is an example of how these two packages work together very well.
combined_age_education_summary <-
  combined_age_education %>% 
  group_by(common_age_category, pop_type) %>% 
  mutate(proportion_of_age = pop_count / sum(pop_count),
         percentage_of_age = scales::percent(proportion_of_age))

# combined_age_education_summary %>% 
#   write.csv(file = "education_attainment_age_prison_vs_general_pop.csv")

# Setting the tibble (fancy data frame) to print all rows/columns
options(tibble.print_max = Inf)
options(tibble.width = Inf)

combined_age_education_summary

```

